# Author: Fabiola Iannarilli
# Date: 2025-01-20

# This script accompains the manuscript:
# "SNAPSHOT EUROPE 2021-2022: The first coordinated, volunteer-based and European-wide camera trap surveys of terrestrial vertebrates"

This code provided in this script contains instructions to clean and collate camera-trap data collected by participants of Snapshot Europe 2021 and 2022. Collaborators uploaded and processed the images collected in September-October of each year in Wildlife Insights. Details on the two datasets and the data collection protocol are available in the metadata of the associated publication @data_paper_2024.

The script is structured as folows:
* __Data cleaning and collating__. We start by cleaning the data of the projects @SE_2021 and @SE_2022 after they have been downloaded from Wildlife Insights. We, then, collate the cleaned version of these two datasets into one, which is the version shared in the associated publication @data_paper_2024. 
* __Data exploration and visualization__. We visually explore the data and create the figures and summaries reported in the metadata file of @data_paper_2024.

## Data cleaning

We load the data and the necessary functions and packages. 

```{r create1}
# Load libraries
library(tidyverse)
#library(terra)
#library(tidyterra)
library(sf)
library(scales)
library(wesanderson)
library(ggpubr)
library(ggh4x)
library(ggnewscale)
source("custom_functions/clean_2021.R")
source("custom_functions/clean_2022.R")
source("custom_functions/collate_dataset.R")

# Load data
# enter date of data download
date_download <- "20250224"

# 2021
se21_proj <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/projects.csv"))
se21_cam <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/cameras.csv"))
se21_depl <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/deployments.csv"))
se21_seq <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/sequences.csv"))
se21_images0 <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/images_2003388_0.csv"))
se21_images1 <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/images_2003388_1.csv"))

# 2022
se22_proj <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/projects.csv"))
se22_cam <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/cameras.csv"))
se22_depl <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/deployments.csv"))
se22_seq <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/sequences.csv"))
se22_images0 <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/images_2004446_0.csv"))
se22_images1 <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/images_2004446_1.csv"))

```

We run test script to fix some known issues in both datasets.

``` {r create2}
cleaned_2021 <- clean_2021(year = 2021)
cleaned_2022 <- clean_2022(year = 2022)
```

We now collate each element (e.g. *seq*, that is the object containing the sequences) of the cleaned dataset for 2021 with the corresponding items of year 2022. 

``` {r create3}
cleaned <- collate_dataset(dat1 = cleaned_2021, dat2 = cleaned_2022) 
```

We save each item in the *cleaned* list in a separate file in a folder named *DataS1*. The overall structure mimics the structure of the relational database used by Wildlife Insights. The zipped version of this folder correspond to the version shared with the associate publication @data_paper_2024. Because the data are presented in the format adopted by Wildlife Insights, we also copy to the output folder their Data Dictionary, which is included in the original data bundles downloaded from the platform.

``` {r create4}
write_csv(cleaned$cam, file = "data_output/DataS1/Snapshot_Europe_2021_2022_cameras.csv")
write_csv(cleaned$depl, file = "data_output/DataS1/Snapshot_Europe_2021_2022_deployments.csv")
write_csv(cleaned$proj, file = "data_output/DataS1/Snapshot_Europe_2021_2022_projects.csv")
write_csv(cleaned$seq, file = "data_output/DataS1/Snapshot_Europe_2021_2022_sequences.csv")
write_csv(cleaned$images, file = "data_output/DataS1/Snapshot_Europe_2021_2022_images.csv")
# move info pdfs
file.copy(from = list.files(paste0("data_input/2021_downloaded_on_", date_download), 
                            pattern = "Dictionary", 
                            full.names = TRUE, 
                            recursive = TRUE), 
          to = "data_output/DataS1/")

```

## Data visualization and mapping

We create a few graphics that capture the spatial, temporal, and taxonomic extent of the data. These visuals are the figures presented in the Metadata document associated to this data paper.

### Spatial coverage
We start with mapping the locations of the arrays in each year in relation to the main biogeoregions occurring in Europe.
``` {r plot1}
# data prep: extract one camera-trap site for each array per year (i.e. subproject)
subproj <- cleaned$depl %>% 
  group_by(year, subproject_name) %>% 
  slice(1) %>% 
  select(year, subproject_name, longitude, latitude, subproject_name_no_year) 

# number of subprojects per year
table(subproj$year)

# was a subproject active in both years?
subproj_both <- subproj %>% 
  group_by(subproject_name_no_year) %>% 
  summarise(n_years = n()) %>% 
  filter(n_years == 2)

# bring info back to loc dataframe
subproj <- subproj %>% 
  mutate(sampled_both = ifelse(subproject_name_no_year %in% subproj_both$subproject_name_no_year, "Yes", "No"))

# Turn off spherical geometry
sf::sf_use_s2(FALSE)

# convert loc to a spatial feature
subproj_sf <- st_as_sf(subproj, coords = c("longitude", "latitude"), crs = 4326)

# load EEA Biogeoregions downloaded from https://www.eea.europa.eu/en/datahub/datahubitem-view/11db8d14-f167-4cd5-9205-95638dfd9618 on 2025-02-18
ecor <- read_sf("other_data/eea_v_3035_1_mio_biogeo-regions_p_2016_v01_r00/BiogeoRegions2016.shp")

# transform projection
subproj_sf <- st_transform(subproj_sf, 
                       crs = st_crs("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs"))
ecor <- st_transform(ecor,
                     crs = st_crs("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs"))

# draw a buffered box around locations
ecor_crop <- st_crop(ecor, st_bbox(subproj_sf) + c(-200000, -200000, 200000, 800000))

# set palette and plot map
pal = wes_palette("Zissou1", n = 11, "continuous")
pl1 <- ggplot() +
  geom_sf(data = ecor_crop, aes(colour = code, fill = code), size = 0.01, alpha = 0.5) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  labs(fill = "Bio-geographical Region", colour = "Bio-geographical Region") +
  guides(colour = guide_legend("Bio-geographical Region"), fill = guide_legend("Bio-geographical Region")) +
  guides(colour = "none", fill = "none") +
  new_scale_colour() +
  new_scale_fill() +
  geom_sf(data = subproj_sf, aes(fill = sampled_both, shape = sampled_both),
          colour = "black", inherit.aes = FALSE, size = 2, alpha = 0.5) +
  scale_fill_manual(values = c("purple", "green")) +
  scale_shape_manual(values = c(21, 24)) +
  labs(fill = "Sampled both years?", shape = "Sampled both years?") +
  guides(colour = guide_legend(ncol = 3, byrow = TRUE)) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom", 
        legend.justification = "top",
        legend.title = element_text(vjust = 0, hjust = 0, face = "bold", size = 12),
        legend.text = element_text(size = 12),
        strip.text = element_text(hjust = 0, face = "bold", size = 14),
        strip.background = element_rect(fill = "white", colour = "white")
        ) +
  facet_wrap(~year, ncol = 2)

# plot number of array by biogeoregion and year
subproj_sf_bgr <- st_intersection(ecor_crop, subproj_sf)
subproj_sf_bgr_summ <- subproj_sf_bgr %>% 
  group_by(code, year) %>% 
  summarise(n_subproj = n())

pl2 <- ggplot(subproj_sf_bgr_summ, aes(x = reorder(code, n_subproj), y = n_subproj, 
                                group = year, color = code, fill = code, alpha = as.factor(year))) +
  geom_bar(stat = "identity", width = 0.7, position = position_dodge(width = 0.8)) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  scale_alpha_manual(values = c(0.8, 0.3), labels = c("2021", "2022")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(fill = "Bio-geographical Region", colour = "Bio-geographical Region", 
       x = "Bio-geographical Region", y = "Number of subprojects") +
  guides(colour = "none", fill = "none") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(face = "bold", size = 12))

# plot effort (i.e. number of sampling days) by biogeoregion and year
effort <- cleaned$depl %>% 
  mutate(n_days = as.numeric(difftime(as.Date(end_date), as.Date(start_date), units = "day"))) %>% 
  group_by(subproject_name) %>% 
  summarise(n_days = sum(n_days))
effort <- subproj_sf_bgr %>% 
  left_join(effort) %>% 
  left_join(as.data.frame(subproj_sf_bgr_summ) %>% select(code, year, n_subproj))


pl3 <- ggplot(effort, aes(x = reorder(code, n_subproj), y = n_days, 
                   color = code, fill = code, alpha = as.factor(year))) +
  geom_boxplot(width = 0.7, position = position_dodge(width = 0.8)) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  scale_alpha_manual(values = c(0.8, 0.3), labels = c("2021", "2022")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_flip() +
  labs(y = "Number of sampling days", x = "Bio-geographical Region") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(face = "bold", size = 12),
        axis.title.y = element_blank(),
        axis.text.y = element_blank())


# bring plots together
ggarrange(pl1, 
          ggarrange(pl2, pl3, ncol = 2),
          nrow = 2, labels = c("A", "B"))
ggarrange(ggarrange(pl2, pl3, ncol = 1), pl1, 
          ncol = 2)
ggsave(filename = "figures/SnapshotEurope2021_and_2022_spatial_coverage_and_effort.png",
       device = "png", units = "cm", height = 29, width = 22, dpi = 600) 

```


### Temporal coverage
We then visualize the period of activity for each array in the two subsequent years.

``` {r plot2}
# data prep: calculate period of activity of each array per year (i.e. subproject)
temp <- cleaned$depl %>%
  group_by(year, subproject_name_no_year) %>% 
  summarize(first_day_depl = min(as.Date(start_date)),
            last_day_depl = max(as.Date(end_date))) %>% 
  mutate(first_day_depl = ymd(first_day_depl),
         last_day_depl = ymd(last_day_depl),
         sampled_both = ifelse(subproject_name_no_year %in% subproj_both$subproject_name_no_year, "Yes", "No")) %>%  
  ungroup()

# count number of location per subproject
loc <- cleaned$depl %>% 
  group_by(year, subproject_name_no_year, placename) %>% 
  distinct() %>% 
  ungroup()%>% 
  group_by(year, subproject_name_no_year) %>% 
  summarize(n_loc = n()) %>% 
  ungroup()

# merge info
temp <- left_join(temp, loc) %>% 
  mutate(n_loc_cut = cut(n_loc, breaks = c(1, 10, 20, 30, 50, 92), include.lowest = TRUE))

# plot
pl_both <- temp %>% 
  filter(sampled_both == "Yes") %>% 
ggplot(., aes(y = subproject_name_no_year, color = n_loc_cut)) +
  geom_segment(aes(x = first_day_depl, xend = last_day_depl,
                   yend = subproject_name_no_year), linewidth = 1) +
  geom_vline(xintercept = as.Date(c("2021-09-01", "2021-10-31",
                                    "2022-09-01", "2022-10-31")), color = "grey80", linetype = "dashed") +
  labs(x = "Sampling period", color = "Number of locations") +
  scale_x_date(labels = date_format("%b")) +
  scale_color_viridis_d() +
  scale_y_discrete(limits=rev) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_line(color = "grey90"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white")) +
  facet_grid(~year, scales = "free", space = "free")

pl_2021_only <- temp %>% 
  filter(sampled_both == "No" & year == "2021") %>% 
ggplot(., aes(y = subproject_name_no_year, color = n_loc_cut)) +
  geom_segment(aes(x = first_day_depl, xend = last_day_depl,
                   yend = subproject_name_no_year), linewidth = 1) +
  geom_vline(xintercept = as.Date(c("2021-09-01", "2021-10-31")), color = "grey80", linetype = "dashed") +
  labs(x = "Sampling period", color = "Number of locations") +
  scale_x_date(labels = date_format("%b")) +
  scale_color_viridis_d() +
  scale_y_discrete(limits=rev) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_line(color = "grey90"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white")) +
  facet_grid(~year, scales = "free", space = "free") 

pl_2022_only <- temp %>% 
  filter(sampled_both == "No" & year == "2022") %>% 
ggplot(., aes(y = subproject_name_no_year, color = n_loc_cut)) +
  geom_segment(aes(x = first_day_depl, xend = last_day_depl,
                   yend = subproject_name_no_year), linewidth = 1) +
  geom_vline(xintercept = as.Date(c("2022-09-01", "2022-10-31")), color = "grey80", linetype = "dashed") +
  labs(x = "Sampling period", color = "Number\nof\nlocations") +
  guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
  scale_x_date(labels = date_format("%b")) +
  scale_color_viridis_d() +
  scale_y_discrete(limits=rev) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 12),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_line(color = "grey90"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white")) +
  facet_grid(~year, scales = "free", space = "free") 


# bring plots together and save
ggarrange(annotate_figure(pl_both, top = text_grob("Sampled in both years", 
                   color = "black", face = "bold", size = 12)), 
          annotate_figure(ggarrange(pl_2021_only, pl_2022_only, ncol = 2), 
                          top = text_grob("Sampled only in year:", color = "black", face = "bold", size = 12)), 
          nrow = 2, common.legend = FALSE)
 
ggsave(filename = "figures/SnapshotEurope2021_and_2022_temporal_coverage.png",
       device = "png", units = "cm", height = 29, width = 22, dpi = 600)
```

### Taxonomic coverage

``` {r plot3}
# species' raw counts
seq <- cleaned$seq %>% 
  filter(species != "" & species != "No CV Result" &
           common_name != "Human-Camera Trapper" &
           common_name != "Human-Researcher" &
           common_name != "Human-Maintenance crew") %>% 
  group_by(year, class, order, family, genus, species) %>% 
  summarise(n_seq = n()) %>% 
  mutate(scientific_name = paste(genus, species, sep = " "))

# plot raw event counts
pl1 <- seq %>% filter(n_seq > 100) %>% 
ggplot(., aes(x = interaction(scientific_name, family, order, class), y = n_seq, 
              group = as.factor(year), color = as.factor(year), fill = as.factor(year))) + 
  geom_bar(stat="identity", alpha = 0.5, position = position_dodge(width = 0.8)) +
  labs(y = "Number of sequences*", x = "", caption = "*Only instances with more than 100 sequences per year") +
  scale_color_manual(values = c("#000000", "#FFBB00")) +
  scale_fill_manual(values = c("#000000", "#FFBB00")) +
  scale_x_discrete(guide = "axis_nested", limits = rev) +
  labs(color = "Year", fill = "Year") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 12),
        axis.title.x = element_text(face = "bold"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white"),
        ggh4x.axis.nestline = element_line(linewidth = 1))

ggsave(pl1, filename = "figures/SnapshotEurope2021_and_2022_taxonomic_coverage_raw_event_counts.png",
       device = "png", units = "cm", height = 29, width = 22, dpi = 600)


# species' detection rates (by subproject)
# species by 
seq2 <- cleaned$seq %>% 
  filter(species != "" & species != "No CV Result" &
           common_name != "Human-Camera Trapper" &
           common_name != "Human-Researcher" &
           common_name != "Human-Maintenance crew") %>% 
  left_join(cleaned$depl %>% select(deployment_id, subproject_name, subproject_name_no_year)) %>% 
  group_by(year, subproject_name, subproject_name_no_year, class, order, family, genus, species) %>% 
  summarise(n_seq = n()) %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  left_join(effort %>% select(year, subproject_name, subproject_name_no_year, n_days)) %>% 
  mutate(det_rates = n_seq/n_days) %>% 
  ungroup()

# plot raw event counts
pl2 <- seq2 %>% filter(scientific_name %in% (seq %>% filter(n_seq > 100))$scientific_name) %>% 
ggplot(., aes(x = interaction(scientific_name, family, order, class), y = log(det_rates*100),
              color = as.factor(year), fill = as.factor(year))) + 
  geom_boxplot(alpha = 0.5, position = position_dodge(width = 0.8)) +
  labs(y = "Detection rates*\n(log10 number of events per 100 trapping days)", x = "", caption = "*Only species with more than 100 sequences per year") +
  scale_x_discrete(guide = "axis_nested", limits = rev) +
  scale_color_manual(values = c("#000000", "#FFBB00")) +
  scale_fill_manual(values = c("#000000", "#FFBB00")) +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none",
        axis.title.x = element_text(face = "bold"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white"),
        ggh4x.axis.nestline = element_line(linewidth = 1))

ggsave(filename = "figures/SnapshotEurope2021_and_2022_taxonomic_coverage_det_rates.png",
       device = "png", units = "cm", height = 29, width = 22, dpi = 600)

# Join plots
ggarrange(pl1 + labs(y = "Number of sequences*\n", caption = ""), pl2 + theme(axis.text.y = element_blank()),
          widths = c(60, 40), ncol = 2, common.legend = TRUE, legend = "top") %>% 
ggsave(filename = "figures/SnapshotEurope2021_and_2022_taxonomic_coverage.png",
       device = "png", units = "cm", height = 22, width = 49, dpi = 600)

```

## Data summaries

``` {r tab1}
# number of locations
table(cleaned$depl$year) 
  
# number of subprojects (i.e. arrays)
cleaned$depl %>% 
  select(year, subproject_name) %>% 
  distinct() %>% 
  group_by(year) %>% 
  summarise(n_subproj = n())

# number of countries
cleaned$depl %>% 
  mutate(country = substr(subproject_name,1,2)) %>% 
  select(year, country) %>% 
  distinct() %>% 
  group_by(year) %>% 
  summarise(n_country = n())

# total number of camera-trap days
sum(temp$tot_effort)

# total number of camera-trap days per year
temp %>% 
  group_by(year) %>% 
  summarise(year_effort = sum(tot_effort))

# % of array sampled both years
length(unique(temp$subproject_name_no_year))
table(table(temp$subproject_name_no_year)>1)

# number of wildlife sequences by class
cleaned$seq %>% 
  filter(is_blank == 0 &
         genus != "Homo" &
           class %in% c("Aves", "Mammalia") &
           species != "") %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  group_by(class) %>% 
  summarise(n_seq = n(),
            n_species = length(unique(scientific_name)))

# period of study
temp %>% 
  group_by(year) %>% 
  summarize(start = min(first_day_depl),
            end = max(last_day_depl))

# camera site per array
n_sites <- cleaned$depl %>% 
  group_by(year, subproject_name) %>% 
  summarize(n_sites = n())
range(n_sites$n_sites)
mean(n_sites$n_sites)
sd(n_sites$n_sites)
```

