# Author: Fabiola Iannarilli
# Date: 2025-01-20

# This script accompains the manuscript:
# "SNAPSHOT EUROPE 2021-2022: The first coordinated, volunteer-based and European-wide camera trap surveys of terrestrial vertebrates"

This code provided in this script contains instructions to clean and collate camera-trap data collected by participants of Snapshot Europe 2021 and 2022. Collaborators uploaded and processed the images collected in September-October of each year in Wildlife Insights. Details on the two datasets and the data collection protocol are available in the metadata of the associated publication @data_paper_2024.

The script is structured as follows:
* __Data cleaning__. We start by cleaning the data of the projects @SE_2021 and @SE_2022 after they have been downloaded from Wildlife Insights. We, then, collate the cleaned version of these two datasets into one, which is the version shared in the associated publication @data_paper_2024. 
* __Data visualization__. We visually explore the data and create the figures reported in the metadata file of @data_paper_2024.
* __Data summaries__. We calculate the summaries that are included in the metadata file of @data_paper_2024.

## Data cleaning

We load the data and the necessary functions and packages. 

```{r create1}
# Load libraries
library(tidyverse)
library(elevatr)
library(sf)
library(scales)
library(wesanderson)
library(ggpubr)
library(ggh4x)
library(ggnewscale)
library(ggtext)
library(glue)
source("custom_functions/clean_2021.R")
source("custom_functions/clean_2022.R")
source("custom_functions/collate_dataset.R")

# Load data
# enter date of data download
date_download <- "20250617"

# 2021
se21_proj <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/projects.csv"))
se21_cam <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/cameras.csv"))
se21_depl <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/deployments.csv"))
se21_seq <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/sequences.csv"))
se21_images0 <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/images_2003388_0.csv"))
se21_images1 <- read.csv(paste0("data_input/2021_downloaded_on_", date_download, "/images_2003388_1.csv"))

# 2022
se22_proj <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/projects.csv"))
se22_cam <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/cameras.csv"))
se22_depl <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/deployments.csv"))
se22_seq <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/sequences.csv"))
se22_images0 <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/images_2004446_0.csv"))
se22_images1 <- read.csv(paste0("data_input/2022_downloaded_on_", date_download, "/images_2004446_1.csv"))

```

We run test script to fix some known issues in both datasets.

``` {r create2}
cleaned_2021 <- clean_2021(year = 2021)
cleaned_2022 <- clean_2022(year = 2022)
```

We now collate each element (e.g. *seq*, that is the object containing the sequences) of the cleaned dataset for 2021 with the corresponding items of year 2022. 

``` {r create3}
cleaned <- collate_dataset(dat1 = cleaned_2021, dat2 = cleaned_2022) 
```

We save each item in the *cleaned* list in a separate file in a folder named *DataS1*. The overall structure mimics the structure of the relational database used by Wildlife Insights. The zipped version of this folder correspond to the version shared with the associate publication @data_paper_2024. 

``` {r create4}
write_csv(cleaned$cam, file = "data_output/DataS1/Snapshot_Europe_2021_2022_cameras.csv")
write_csv(cleaned$depl, file = "data_output/DataS1/Snapshot_Europe_2021_2022_deployments.csv")
write_csv(cleaned$proj, file = "data_output/DataS1/Snapshot_Europe_2021_2022_projects.csv")
write_csv(cleaned$seq, file = "data_output/DataS1/Snapshot_Europe_2021_2022_sequences.csv")
write_csv(cleaned$images, file = "data_output/DataS1/Snapshot_Europe_2021_2022_images.csv")

```

## Data visualization and mapping

We create a few graphics that capture the spatial, temporal, and taxonomic extent of the data. These visuals are the figures presented in the Metadata document associated to this data paper.

### Spatial coverage
We start with mapping the locations of the arrays in each year in relation to the main biogeoregions occurring in Europe.
``` {r plot1}
# data prep: extract one camera-trap site for each array per year (i.e. subproject)
subproj <- cleaned$depl %>% 
  group_by(year, subproject_name) %>% 
  slice(1) %>% 
  select(year, subproject_name, longitude, latitude, subproject_name_no_year) 

# number of subprojects per year
table(subproj$year)

# was a subproject active in both years?
subproj_both <- subproj %>% 
  group_by(subproject_name_no_year) %>% 
  summarise(n_years = n()) %>% 
  filter(n_years == 2)

# bring info back to subproj dataframe
subproj <- subproj %>% 
  mutate(sampled_both = ifelse(subproject_name_no_year %in% subproj_both$subproject_name_no_year, "Yes", "No"))

# Turn off spherical geometry
sf::sf_use_s2(FALSE)

# convert loc to a spatial feature
subproj_sf <- st_as_sf(subproj, coords = c("longitude", "latitude"), crs = 4326)

# load EEA Biogeoregions downloaded from https://www.eea.europa.eu/en/datahub/datahubitem-view/11db8d14-f167-4cd5-9205-95638dfd9618 on 2025-02-18
ecor <- read_sf("other_data/eea_v_3035_1_mio_biogeo-regions_p_2016_v01_r00/BiogeoRegions2016.shp")

# transform projection
subproj_sf <- st_transform(subproj_sf, 
                       crs = st_crs("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs"))
ecor <- st_transform(ecor,
                     crs = st_crs("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs"))

# draw a buffered box around locations
ecor_crop <- st_crop(ecor, st_bbox(subproj_sf) + c(-200000, -200000, 200000, 800000))

# prep for mapping
subproj_sf2 <- subproj_sf %>% mutate(sampled = ifelse(sampled_both == "Yes", "Both", as.character(year))) %>% group_by(subproject_name_no_year, sampled) %>% slice(1)

# set palette and plot map
pal = wes_palette("Zissou1", n = 11, "continuous")
pl1 <- ggplot() +
    geom_sf(data = ecor_crop, aes(colour = code, fill = code), size = 0.01, alpha = 0.5) +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(fill = "Bio-geographical Region", colour = "Bio-geographical Region") +
    guides(colour = guide_legend("Bio-geographical Region"), fill = guide_legend("Bio-geographical Region")) +
    new_scale_colour() +
    new_scale_fill() +
    geom_sf(data = subproj_sf2, aes(fill = sampled, shape = sampled),
            colour = "black", inherit.aes = FALSE, size = 2, alpha = 0.8) +
    scale_fill_manual(values = c("purple", "green", "blue")) +
    scale_shape_manual(values = c(21, 24, 22)) +
    labs(fill = "Year(s) sampled", shape = "Year(s) sampled") +
    guides(colour = guide_legend(ncol = 3, byrow = TRUE)) +
    ggthemes::theme_map() +
    theme(legend.position = "right", 
          legend.justification = "top",
          legend.title = element_text(vjust = 0, hjust = 0, face = "bold", size = 12),
          legend.text = element_text(size = 12),
          strip.text = element_text(hjust = 0, face = "bold", size = 14),
          strip.background = element_rect(fill = "white", colour = "white")
    )

# plot number of array by biogeoregion and year
subproj_sf_bgr <- st_intersection(ecor_crop, subproj_sf)
subproj_sf_bgr_summ <- subproj_sf_bgr %>% 
  group_by(code, year) %>% 
  summarise(n_subproj = n()) %>% 
  ungroup() %>% 
  mutate(year_ft = factor(year, levels = c("2022", "2021")))

pl2 <- ggplot(subproj_sf_bgr_summ, aes(x = reorder(code, n_subproj), y = n_subproj, 
                                group = year_ft, color = code, fill = code, alpha = year_ft)) +
  geom_bar(stat = "identity", width = 0.7, position = position_dodge(width = 0.8)) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  scale_alpha_manual(values = c(0.3, 0.8), labels = c("2022", "2021")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(fill = "Bio-geographical Region", colour = "Bio-geographical Region", 
       x = "Bio-geographical Region", y = "Number of subprojects") +
  guides(colour = "none", fill = "none") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(face = "bold", size = 12))

# plot effort (i.e. number of sampling days) by biogeoregion and year
effort <- cleaned$depl %>% 
  mutate(n_days = as.numeric(difftime(as.Date(end_date), as.Date(start_date), units = "day"))) %>% 
  group_by(subproject_name) %>% 
  summarise(n_days = sum(n_days),
            n_loc = length(unique(placename)))
effort <- subproj_sf_bgr %>% 
  left_join(effort) %>% 
  left_join(as.data.frame(subproj_sf_bgr_summ) %>% select(code, year, n_subproj)) %>% 
  mutate(year_ft = factor(year, levels = c("2022", "2021")))


pl3 <- ggplot(effort, aes(x = reorder(code, n_subproj), y = n_days, 
                   color = code, fill = code, alpha = year_ft)) +
  geom_boxplot(width = 0.7, position = position_dodge(width = 0.8)) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  scale_alpha_manual(values = c(0.3, 0.8), labels = c("2022", "2021")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_flip() +
  labs(y = "Number of sampling days", x = "Bio-geographical Region") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(face = "bold", size = 12),
        axis.title.y = element_blank(),
        axis.text.y = element_blank())


# bring plots together and save
ggarrange(pl1, 
          ggarrange(pl2, pl3, ncol = 2),
          nrow = 2, labels = c("A", "B")) %>% 
ggsave(filename = "figures/SnapshotEurope2021_and_2022_spatial_coverage_and_effort.png",
       device = "png", units = "cm", height = 29, width = 22, dpi = 600) 

```


### Temporal coverage
We then visualize the period of activity for each array in the two subsequent years.

``` {r plot2}
# data prep: calculate period of activity of each array per year (i.e. subproject)
temp <- cleaned$depl %>%
  group_by(year, subproject_name_no_year) %>% 
  summarize(first_day_depl = min(as.Date(start_date)),
            last_day_depl = max(as.Date(end_date))) %>% 
  mutate(first_day_depl = ymd(first_day_depl),
         last_day_depl = ymd(last_day_depl),
         sampled_both = ifelse(subproject_name_no_year %in% subproj_both$subproject_name_no_year, "Yes", "No")) %>%  
  ungroup()

# count number of location per subproject
loc <- cleaned$depl %>% 
  group_by(year, subproject_name_no_year, placename) %>% 
  distinct() %>% 
  ungroup()%>% 
  group_by(year, subproject_name_no_year) %>% 
  summarize(n_loc = n()) %>% 
  ungroup()

# merge info
temp <- left_join(temp, loc) %>% 
  mutate(n_loc_cut = cut(n_loc, breaks = c(1, 10, 20, 30, 50, 92), include.lowest = TRUE))

# plot
pl_both <- temp %>% 
  filter(sampled_both == "Yes") %>% 
ggplot(., aes(y = subproject_name_no_year, color = n_loc_cut)) +
  geom_segment(aes(x = first_day_depl, xend = last_day_depl,
                   yend = subproject_name_no_year), linewidth = 1) +
  geom_vline(xintercept = as.Date(c("2021-09-01", "2021-10-31",
                                    "2022-09-01", "2022-10-31")), color = "grey80", linetype = "dashed") +
  labs(x = "Sampling period", color = "Number of locations") +
  scale_x_date(labels = date_format("%b")) +
  scale_color_viridis_d() +
  scale_y_discrete(limits=rev) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_line(color = "grey90"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white")) +
  facet_grid(~year, scales = "free", space = "free")

pl_2021_only <- temp %>% 
  filter(sampled_both == "No" & year == "2021") %>% 
ggplot(., aes(y = subproject_name_no_year, color = n_loc_cut)) +
  geom_segment(aes(x = first_day_depl, xend = last_day_depl,
                   yend = subproject_name_no_year), linewidth = 1) +
  geom_vline(xintercept = as.Date(c("2021-09-01", "2021-10-31")), color = "grey80", linetype = "dashed") +
  labs(x = "Sampling period", color = "Number of locations") +
  scale_x_date(labels = date_format("%b")) +
  scale_color_viridis_d() +
  scale_y_discrete(limits=rev) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_line(color = "grey90"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white")) +
  facet_grid(~year, scales = "free", space = "free") 

pl_2022_only <- temp %>% 
  filter(sampled_both == "No" & year == "2022") %>% 
ggplot(., aes(y = subproject_name_no_year, color = n_loc_cut)) +
  geom_segment(aes(x = first_day_depl, xend = last_day_depl,
                   yend = subproject_name_no_year), linewidth = 1) +
  geom_vline(xintercept = as.Date(c("2022-09-01", "2022-10-31")), color = "grey80", linetype = "dashed") +
  labs(x = "Sampling period", color = "Number\nof\nlocations") +
  guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
  scale_x_date(labels = date_format("%b")) +
  scale_color_viridis_d() +
  scale_y_discrete(limits=rev) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 12),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_line(color = "grey90"),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white")) +
  facet_grid(~year, scales = "free", space = "free") 


# bring plots together and save
ggarrange(annotate_figure(pl_both, top = text_grob("Sampled in both years", 
                   color = "black", face = "bold", size = 12)), 
          annotate_figure(ggarrange(pl_2021_only, pl_2022_only, ncol = 2), 
                          top = text_grob("Sampled only in year:", color = "black", face = "bold", size = 12)), 
          nrow = 2, common.legend = FALSE) %>% 
ggsave(filename = "figures/SnapshotEurope2021_and_2022_temporal_coverage.png",
       device = "png", units = "cm", height = 29, width = 22, dpi = 600)
```

### Taxonomic coverage

``` {r plot3}
# species' raw counts
seq <- cleaned$seq %>% 
  filter(species != "" & species != "No CV Result" &
           common_name != "Human-Camera Trapper" &
           common_name != "Human-Researcher" &
           common_name != "Human-Maintenance crew") %>% 
  group_by(year, class, order, family, genus, species) %>% 
  summarise(n_seq = n()) %>% 
  mutate(scientific_name = paste(genus, species, sep = " "),
         year_ft = factor(year, levels = c("2022", "2021")))

# plot raw event counts
pl_taxo1 <- seq %>% filter(n_seq > 100) %>% 
  mutate(scientific_name_ita = glue("<i>{scientific_name}</i>")) %>%  
  ggplot(., aes(x = interaction(scientific_name_ita, family, order, class), y = n_seq, 
                group = year_ft, color = year_ft, fill = year_ft)) + 
  geom_bar(stat="identity", alpha = 0.5, position = position_dodge(width = 0.8)) +
  labs(y = "Number of sequences*", x = "", caption = "*Only instances with more than 100 sequences per year") +
  scale_color_manual(values = c("#FFBB00", "#000000")) +
  scale_fill_manual(values = c("#FFBB00", "#000000")) +
  scale_x_discrete(guide = "axis_nested", limits = rev) +
  labs(color = "Year", fill = "Year") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none",
        legend.title = element_text(face = "bold", size = 12),
        axis.title.x = element_text(face = "bold"),
        axis.text.y = element_markdown(size = 12),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white"),
        ggh4x.axis.nestline = element_line(linewidth = 1))

# species' detection rates (by subproject)
# species by 
seq2 <- cleaned$seq %>% 
  filter(species != "" & species != "No CV Result" &
           common_name != "Human-Camera Trapper" &
           common_name != "Human-Researcher" &
           common_name != "Human-Maintenance crew") %>% 
  left_join(cleaned$depl %>% select(deployment_id, subproject_name, subproject_name_no_year)) %>% 
  group_by(year, subproject_name, subproject_name_no_year, class, order, family, genus, species) %>% 
  summarise(n_seq = n()) %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  left_join(effort %>% select(year, subproject_name, subproject_name_no_year, n_days)) %>% 
  mutate(det_rates = n_seq/n_days) %>% 
  ungroup() %>% 
  mutate(year_ft = factor(year, levels = c("2022", "2021")))

# plot raw event counts
pl_taxo2 <- seq2 %>% filter(scientific_name %in% (seq %>% filter(n_seq > 100))$scientific_name) %>% 
  mutate(scientific_name_ita = glue("<i>{scientific_name}</i>")) %>% 
  ggplot(., aes(x = interaction(scientific_name_ita, family, order, class), y = log(det_rates*100),
              color = year_ft, fill = year_ft)) + 
  geom_boxplot(alpha = 0.5, position = position_dodge(width = 0.8)) +
  labs(y = "Detection rates*\n(log10 number of events per 100 trapping days)", x = "", caption = "*Only species with more than 100 sequences per year") +
  scale_x_discrete(guide = "axis_nested", limits = rev) +
  scale_color_manual(values = c("#FFBB00", "#000000")) +
  scale_fill_manual(values = c("#FFBB00", "#000000")) +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none",
        axis.title.x = element_text(face = "bold"),
        axis.text.y = element_markdown(size = 12),
        strip.text = element_text(hjust = 0, face = "bold"),
        strip.background = element_rect(fill = "white", colour = "white"),
        ggh4x.axis.nestline = element_line(linewidth = 1))

# Join plots
ggarrange(pl_taxo1 + labs(caption = ""), pl_taxo2, 
          heights = c(50, 50), ncol = 1, common.legend = FALSE) %>% 
ggsave(filename = "figures/SnapshotEurope2021_and_2022_taxonomic_coverage_vertical.png",
       device = "png", units = "cm", height = 34, width = 22, dpi = 600)

```

## Data summaries

``` {r summaries}
# number of locations
cleaned$depl %>% 
  select(project_id, placename, year) %>% 
  distinct() %>% 
  group_by(year) %>% 
  summarise(n_locs = n())

# number of deployments
table(cleaned$depl$year) 
  
# number of subprojects (i.e. arrays)
cleaned$depl %>% 
  select(year, subproject_name) %>% 
  distinct() %>% 
  group_by(year) %>% 
  summarise(n_subproj = n())

# number of countries
cleaned$depl %>% 
  mutate(country = substr(subproject_name,1,2)) %>% 
  select(year, country) %>% 
  distinct() %>% 
  group_by(year) %>% 
  summarise(n_country = n())

# total number of camera-trap days
sum(effort$n_days)

# total number of camera-trap days per year
effort %>% 
  group_by(year) %>% 
  summarise(year_effort = sum(n_days))

# % of array sampled both years
length(unique(effort$subproject_name_no_year))
table(table(effort$subproject_name_no_year)>1)
100*table(table(effort$subproject_name_no_year)>1)/length(unique(effort$subproject_name_no_year))

# number of wildlife sequences by class
cleaned$seq %>% 
  filter(is_blank == 0 &
         genus != "Homo" &
           class %in% c("Aves", "Mammalia") &
           species != "") %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  group_by(class) %>% 
  summarise(n_seq = n(),
            n_species = length(unique(scientific_name)))

# period of study
temp %>% 
  group_by(year) %>% 
  summarize(start = min(first_day_depl),
            end = max(last_day_depl))

# camera site per array
n_sites <- cleaned$depl %>% 
  group_by(year, subproject_name) %>% 
  summarize(n_sites = n())
range(n_sites$n_sites)
mean(n_sites$n_sites)
sd(n_sites$n_sites)

# trap days
range(effort$n_days)
mean(effort$n_days)
sd(effort$n_days)

# number of sequences per species
cleaned$seq %>% 
    filter(is_blank == 0 &
               genus != "Homo" &
               class %in% c("Aves", "Mammalia") &
               species != "") %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  group_by(class, scientific_name) %>% 
  summarise(n_seq = n())

# number of sequences per species and year
cleaned$seq %>% 
    filter(is_blank == 0 &
               genus != "Homo" &
               class %in% c("Aves", "Mammalia") &
               species != "") %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  group_by(class, scientific_name, year) %>% 
  summarise(n_seq = n())

# number of sequences per Homo sapiens (excluding research activities)
table((cleaned$seq %>% 
    filter(is_blank == 0 &
               genus == "Homo" &
           common_name != "Human-Camera Trapper" &
           common_name != "Human-Researcher" &
           common_name != "Human-Maintenance crew") %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")))$year)

# number of sequences per domestic animals
table((cleaned$seq %>% 
    filter(is_blank == 0) %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  filter(scientific_name %in% c("Canis familiaris", "Felis catus", "Bos taurus", 
                                "Equus asinus", "Equus caballus", "Gallus gallus domesticus",
                                "Ovis aries")))$year)

# latitudinal and longitudinal range by year
subproj %>% 
  group_by(year) %>% 
  summarise(min_lat = min(latitude),
            max_lat = max(latitude),
            min_lng = min(longitude),
            max_lng = max(longitude))

# range in elevation
ele <- get_elev_point(subproj_sf, src = "aws")
ele %>% 
  group_by(year) %>% 
  summarise(min_ele = min(elevation),
            max_ele = max(elevation))

# percent sequences outside Sept-Oct
table((cleaned$seq %>% 
  mutate(inside = ifelse(month(start_time) %in% c("9", "10"), "Yes", "No")))$inside)

# habitat types
subproj %>% 
  mutate(habitat = str_split(subproject_name, pattern = "_")[[1]][2]) %>% 
  group_by(year, habitat) %>% 
  summarise(n_subproj = n())

# files' dimension and size
lapply(cleaned, dim)
files <- list.files("data_output/DataS1/", full.names = TRUE)
(file.info(files))$size/1000
```

## Tables
The following code creates Tables 1 and 2 reported in the Metadata document associated to this data paper.

``` {r tables}
# Table 1. Sampling effort.
tab1 <- as.data.frame(effort) %>%
  select(year, subproject_name_no_year, n_loc, n_days) %>% 
  pivot_wider(id_cols = subproject_name_no_year, names_from = year, values_from = c("n_days", "n_loc"), values_fill = 0)
table(tab1$n_days_2021)
table(tab1$n_days_2022)
table(tab1$n_loc_2021)
table(tab1$n_loc_2022)
write.csv(tab1, file = "tables/Table_Effort_by_subproject.csv")  

# Table 2. Species list.
# number of sequences per species and year
wild <- cleaned$seq %>% 
    filter(is_blank == 0 &
               genus != "Homo" &
               class %in% c("Aves", "Mammalia") &
               species != "") %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  filter(!(scientific_name %in% c("Canis familiaris", "Felis catus", "Bos taurus", 
                                "Equus asinus", "Equus caballus", "Gallus gallus domesticus",
                                "Ovis aries", "Sus scrofa scrofa"))) %>% 
  group_by(class, scientific_name, common_name, year) %>% 
  summarise(n_seq = n()) %>% 
  mutate(Group = "Wild species") %>% 
  pivot_wider(id_cols = NULL, names_from = year, values_from = n_seq, values_fill = 0) %>% 
  select(Group, class, common_name, scientific_name, `2021`, `2022`) 

# number of sequences per Homo sapiens (excluding research activities)
homo <- cleaned$seq %>% 
    filter(is_blank == 0 &
               genus == "Homo" &
           common_name != "Human-Camera Trapper" &
           common_name != "Human-Researcher" &
           common_name != "Human-Maintenance crew") %>% 
  mutate(scientific_name = "Homo sapiens",
         common_name = "Human") %>% 
  group_by(class, scientific_name, common_name, year) %>% 
  summarise(n_seq = n()) %>% 
  mutate(Group = "Human") %>% 
  select(Group, class, common_name, scientific_name, year, n_seq) %>% 
  pivot_wider(id_cols = NULL, names_from = year, values_from = n_seq, values_fill = 0)
  

# number of sequences per domestic animals
domestic <- cleaned$seq %>% 
    filter(is_blank == 0) %>% 
  mutate(scientific_name = paste(genus, species, sep = " ")) %>% 
  filter(scientific_name %in% c("Canis familiaris", "Felis catus", "Bos taurus", 
                                "Equus asinus", "Equus caballus", "Gallus gallus domesticus",
                                "Ovis aries", "Sus scrofa scrofa")) %>% 
  mutate(common_name = ifelse(common_name == "Dog-on-leash", "Domestic Dog", common_name)) %>% 
  group_by(class, scientific_name, common_name, year) %>% 
  summarise(n_seq = n()) %>% 
  mutate(Group = "Domestic species") %>% 
  select(Group, class, common_name, scientific_name, year, n_seq) %>% 
  pivot_wider(id_cols = NULL, names_from = year, values_from = n_seq, values_fill = 0)

tab2 <- rbind(wild, domestic, homo)
write.csv(tab2, file = "tables/Table_Sequences_by_species.csv")  

```